Parsimonious Representation of Nonlinear Dynamical Systems Through Manifold Learning: A Chemotaxis Case Study

### 2015-03-12 chemotaxis/paper/draft1.tex
updated paper

\documentclass[preprint]{elsarticle}

@@

\title{Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study}

\author[princeton]{Carmeline~J.~Dsilva\corref{cor1}}
\ead{cdsilva@princeton.edu}
%
\author[technion]{Ronen~Talmon}
\ead{ ronen@ee.technion.ac.il}
%
\author[yale]{Ronald R. Coifman}
\ead{coifman@math.yale.edu}
%
\author[princeton, princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
\ead{yannis@princeton.edu}
%
\address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
\address[technion]{Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
\address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
\address[princetonpacm]{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08540, USA}
%
\cortext[cor1]{Corresponding author}

\begin{abstract}
Nonlinear manifold learning algorithms have proven useful for the anlaysis of complex data sets. 
%
However, these algorithms still encounter several challenges when faced with real data. 
\end{abstract}

@@

\begin{keyword}
TODO
\end{keyword}

@@

In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{....}. 
%
In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system.
%
For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
%
For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies give rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.

%For many engineering applications, it is often advantageous or even essential to effectively parameterize a system's dynamics. 
%TODO: add reasons why, potential applications, etc.
%%
%Often, such a parameterization cannot be obtained analytically, due to the complexity of the system.
%%
%For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
%%
%In such cases, we utilize data obtained from observations and/or simulations of the dynamical system, and turn to data-driven analysis methods to provide an accurate low-dimensional description of this data.
%%
%This low-dimensional description can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior. 
%

Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a nonlinear data-driven technique, to analyze data.
%
Most manifold learning algorithms \cite{...} construct parameterizations of the data through the spectral analysis of a Laplace operator.
%
The data is then embedded in a new, low-dimensional coordinate system given by the eigenvectors of this operator. 
%
Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale {\em ab initio} model.

These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility. 
%
Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{...}. 
%
A critical shortcoming of these methods is that from a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are ``repeated eigendirections'' which describe the same mode. 
%
Identifying those eigenvectors is critical for obtaining a parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
%
This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{...}.

In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression.
%
We first demonstrate our algorithm on synthetic examples where we have a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator, which will allow us to validate our proposed approach.
%
We then consider a complex data set from a stochastic dynamical system which models cellular chemotaxis. 
%
The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which can successfully analyze data from this simulation.
%
This pipeline will allow us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.

%\begin{itemize}
%
%\item
%Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc). 
%
%\item
%Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.
%
%\item
%Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").
%
%\item
%Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome"). 
%
%\item
%However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.
%
%\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density
%
%\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections. 
%
%\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example. The main idea is the ability to obtain, in a data driven manner, the "true" parsimonious description of the system, which is certain cases (such as particular regimes/modes of the chemotaxis system) may be significantly different from the typical representation and choice of variables used to describe the system (e.g. in the chemotaxis, it will not necessarily be the two fluxes).  
%
%\item
%In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.
%
%\item
%All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline. We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).
%
%\end{itemize}

@@

Although data from dynamical systems is most often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
%
In such cases, the high-dimensional microscale data will be restricted to a low-dimensional manifold embedded in the high-dimensional space. 
%
Extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics \cite{...}.

Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$. 
%
We will show how to extract a $d$-dimensional parameterization of the observations which is consistent with the geometry of the manifold.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_cnts2}
\caption{}
\end{subfigure}
\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ .}
\label{fig:strip_efuncs}
\end{figure}

We start by considering a continuous setting where rigorous analysis is available for specific examples. 
%
In such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial. 
%
The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis. 
%
Thus, we consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
%
For linear hyperplanes, the principal axes parameterize the manifold.
%
However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent.

In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
%
To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}). 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper. 
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any additional directions along the strip; we will refer to these as ``repeated eigendirections,'' and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
We note that, although the eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_discrete4}
\caption{}
\end{subfigure}

\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\caption{}
\end{subfigure}
\caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps. (c, d) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps. Note that in both cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$. }
\label{fig:strip_evecs}
\end{figure}

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator. 
%
It was shown in \cite{...} that, in the limit of infinite data, the discrete Laplacian constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold. 
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator \cite{...}.

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}
%

However, the data need not be sampled uniformly from the underlying manifold.
%
If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then the matrix $I-A$ approximates the Fokker-Planck operator, 
\begin{equation}
(I-A) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, 
\end{equation}
where $U = -2 \log q$. 
%
Alternatively, one can factor out the density effects in the weight matrix.
%
Given data $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix
%
\begin{equation}
\tilde{W} = D^{-1} W D^{-1},
\end{equation}
%
where pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
%
We then construct the diagonal matrix $\tilde{D} \in \mathbb{R}^{m \times m}$, with $\tilde{D}_{ii} = \sum_j \tilde{W}_{ij}$, and the matrix $\tilde{A}  = \tilde{D}^{-1} \tilde{W}.$
%
The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator on $\mathcal{M}_d$ even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.

As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{...}.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.%, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
%
Figure~\ref{fig:strip_evecs} shows data sampled from a strip, colored by eigenvectors of $A$. 
%
In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
%
Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$ provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.  
%
From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these eigenvectors are guaranteed to appear before any coupled or repeated eigendirections. 
%
However, these two eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
In practice, identifying which eigenvectors correspond to unique eigendirections is nontrivial, as the correspondence between the original data and the coordinates which parameterize the manifold are unknown.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold cannot easily be obtained from the coordinates of the original data.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\end{subfigure}
%
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_discrete2}
\end{subfigure}

\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_discrete3}
\end{subfigure}
%
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{strip_discrete4}
\end{subfigure}
\caption{ Two-dimensional strip with uniform sampling colored by the first four (non-trivial) eigenvectors from diffusion maps. Note that the first and fourth eigenvectors are one-to-one with $z_1$ and $z_2$, respectively. However, the second and third eigenvectors are higher harmonics of the first eigenvector and do not capture any additional structure within the data set. }
\label{fig:strip_harmonics}
\end{figure}

>>>>

\documentclass[3p]{elsarticle}

@@

\title{Parsimonious Representation of Nonlinear Dynamical Systems Through Manifold Learning: A Chemotaxis Case Study}

\author[princeton]{Carmeline~J.~Dsilva}
\ead{cdsilva@princeton.edu}
%
\author[technion]{Ronen~Talmon}
\ead{ ronen@ee.technion.ac.il}
%
\author[yale]{Ronald R. Coifman}
\ead{coifman@math.yale.edu}
%
\author[princeton,princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
\ead{yannis@princeton.edu}

\address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
\address[technion]{Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
\address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
\address[princetonpacm]{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08540, USA}
%
\cortext[cor1]{Corresponding author}

\begin{abstract}
Nonlinear manifold learning algorithms have been effective for the analysis of complex data sets. 
%
However, these algorithms still encounter several challenges when faced with real data.
%
A key shortcoming is the existence of repeated eigendirections, where several embedding coordinates parameterize the same direction in the data set, therefore obscuring the true dimensionality of the underlying manifold. 
%
In this paper, we propose an algorithm, based on local linear regression, to automatically detect which coordinates are repeated eigendirections. 
%
We first demonstrate the utility and flexibility of our approach on synthetic data sets. 
%
We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis. 
%
Here, we show that our approach for factoring out repeated eigendirections can allow us to detect changes in dynamical behavior and the underlying system dimensionality directly from data. 
%
\end{abstract}

@@

\begin{keyword}
diffusion maps, repeated eigendirections, chemotaxis
\end{keyword}

@@

In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{gepshtein2013image, fernandez2014diffusion, singer2011viewing, yuan2014automated, zhao2003face, trapnell2014dynamics, kemelmacher2011exploring, sifre2013rotation}. 
%
In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system \cite{talmon2014intrinsic,berry2013time,singer2009detecting,ferguson2010systematic}.
%
For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
%
For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies gives rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.

Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
%
Most manifold learning algorithms construct parameterizations of the data through the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
%
The data is then embedded in a new, low-dimensional coordinate system given by the eigenvectors of this operator. 
%
Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale {\em ab initio} model.

These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}. 
%
Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally,rohrdanz2011determination.}.
%
A critical shortcoming of these methods is that from a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are ``repeated eigendirections'' which describe the same mode. 
%
Identifying those eigenvectors is critical for obtaining a parsimonious parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
%
This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.

In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression \cite{wasserman2006all}.
%
We first demonstrate our algorithm on synthetic examples where a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator is known, which will allow us to validate our proposed approach.
%
We then consider a complex data set from a stochastic dynamical system simulation which models cellular chemotaxis \cite{othmer1988models}. 
%
The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which successfully analyzes data from this simulation.
%
This pipeline allows us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.

@@

Although data from dynamical systems is often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
%
In such cases, the high-dimensional microscale data is restricted to a low-dimensional manifold embedded in the high-dimensional space, and extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics.
%
Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$. 
%
We will show how to extract a $d$-dimensional parameterization of the observations which is consistent with the geometry of the manifold.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\includegraphics[width=\textwidth]{strip_cnts2}
\caption{}
\label{subfig:strip_efuncs}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\includegraphics[width=\textwidth]{strip_discrete4}
\caption{}
\label{subfig:strip_evecs_uniform}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\caption{}
\label{subfig:strip_evecs_nonuniform}
\end{subfigure}
%
\caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$. (b) Two-dimensional strip with uniform sampling colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian. (c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian. Note that in all cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.}
\end{figure}

We start by considering a continuous setting where rigorous analysis is available for specific examples. 
%
Using such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial. 
%
The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis.

We consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
%
For linear hyperplanes, the principal axes parameterize the manifold.
%
However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent. 
%
In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
%
To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}). 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper. 
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any additional directions along the strip; we will refer to these as ``repeated eigendirections,'' and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
We note that, although the eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.

@@

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator. 
%
It was shown in \cite{coifman2006geometric} that, in the limit of infinite data, the discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold. 
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator.

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the discrete Laplacian matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}

However, the data need not be sampled uniformly from the underlying manifold.
%
If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then the matrix $I-A$ approximates the Fokker-Planck operator, 
$%\begin{equation}
(I-A) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, 
$ %\end{equation}
where $U = -2 \log q$. 
%
Alternatively, one can factor out the density effects in the weight matrix.
%
Given data $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix
%
\begin{equation}
\tilde{W} = D^{-1} W D^{-1},
\end{equation}
%
where pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
%
We then construct the diagonal matrix $\tilde{D} \in \mathbb{R}^{m \times m}$, with $\tilde{D}_{ii} = \sum_j \tilde{W}_{ij}$, and the matrix $\tilde{A}  = \tilde{D}^{-1} \tilde{W}.$
%
The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator on $\mathcal{M}_d$ even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.

As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{coifman2005geometric, coifman2006geometric}.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
%
Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows data sampled from a strip, colored by eigenvectors of $A$. 
%
In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
%
Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$ provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.  
%
From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these two eigenvectors are guaranteed to appear before any coupled or repeated eigendirections. 
%
However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, and identifying which eigenvectors correspond to unique eigendirections is nontrivial.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete2}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete3}
\end{subfigure}
%
\begin{subfigure}{0.24\textwidth}
\includegraphics[width=\textwidth]{strip_discrete4}
\end{subfigure}
\caption{ Two-dimensional strip with uniform sampling colored by the first four (non-trivial) eigenvectors from diffusion maps. Note that the first and fourth eigenvectors are one-to-one with $z_1$ and $z_2$, respectively. However, the second and third eigenvectors are higher harmonics of the first eigenvector and do not capture any additional structure within the data set. }
\label{fig:strip_harmonics}
\end{figure}
